---
title: "progenitor-score: a worked example on simulated data (R)"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4.6, dpi = 110)
```

`progenitor-score` scores every cell against the **BoneMarrowMap** atlas (Zeng et al., *Blood Cancer Discovery* 2025;6:307-24) and returns the best-matching bone-marrow state, a group, and a call that says whether a cell is a progenitor you may want to keep (for example when extracting monocytes, macrophages and DCs from a public dataset).

This report uses **simulated counts with known labels** (`examples/data/`, made by `examples/simulate_counts.py`). It shows how to call the tool from R, on a plain matrix and on a Seurat object, and how the calls behave on clean, transitional, out-of-reference and low-depth cells. **It is a plumbing and behaviour check, not an accuracy estimate**: the simulated profiles are averages of the atlas the model was trained on, so real cells will do worse. Accuracy on 9 held-out real donors is in `docs/validation/`. The Python notebook (`progenitor_score_demo.ipynb`) runs the same steps; section 6 checks that the two give identical answers.

Requirements: **raw counts** (not normalised), gene **symbols** as row names, ideally a QC flag per cell.

## 1. Load the package and the data

```{r load}
library(Matrix); library(ggplot2)
if (requireNamespace("progenitorscore", quietly = TRUE)) {            # installed:  remotes::install_github("finlay-immuno-lab/progenitor-score")
  library(progenitorscore)
} else {
  source("../R/progenitor_score.R")                                  # or run from a repository checkout
}
counts <- readMM(gzfile("data/sim_counts.mtx.gz"))                    # genes x cells, sparse
genes  <- readLines("data/sim_genes.tsv"); cells <- read.csv("data/sim_cells.csv", stringsAsFactors = FALSE)
cells$qc_pass <- as.logical(cells$qc_pass)                            # written by Python as "True"/"False"
dimnames(counts) <- list(genes, cells$cell)
counts <- as(counts, "CsparseMatrix")
cat(sprintf("%d genes x %d cells; integer counts: %s; max count: %d\n", nrow(counts), ncol(counts),
            all(counts@x == round(counts@x)), max(counts@x)))
head(cells[, c("cell", "true_population", "umis", "qc_pass")], 4)
```

## 2. Score a matrix

`bmps_score()` takes a genes x cells matrix (dense or sparse) with symbols as row names and returns one row per cell. `qc` is a logical vector; cells that are `FALSE` are not scored.

```{r score}
res <- bmps_score(counts, qc = cells$qc_pass, quiet = TRUE)            # defaults: min_margin = 100, min_top1 = 0
head(cbind(true_population = cells$true_population, res), 6)
```

## 3. Score a Seurat object

`bmps_score_seurat()` reads the `counts` layer of the default assay and adds the columns, prefixed `bmps_`, to the object metadata. With Seurat v5 and split layers, run `JoinLayers()` first. If row names are Ensembl ids, pass `symbols = ` a vector of gene symbols.

```{r seurat}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
  obj <- SeuratObject::CreateSeuratObject(counts = counts, meta.data = data.frame(qc_pass = cells$qc_pass, row.names = cells$cell), min.cells = 0, min.features = 0)
  obj <- bmps_score_seurat(obj, qc_col = "qc_pass")
  print(head(obj[[]][, c("qc_pass", "bmps_bm_state", "bmps_margin", "bmps_progenitor_call")], 4))
  stopifnot(identical(unname(obj$bmps_progenitor_call), res$progenitor_call))
  cat("Seurat result == matrix result\n")
}
```

## 4. What do the calls look like for each simulated population?

Rows are the true populations (never shown to the tool). `progenitor_in` = GMP-like or DC precursor with a clear margin; `progenitor_review` = HSC/MPP, LMPP, MLP, neutrophil-primed GMP; `monocyte_precursor` = ProMono stages (part of the monocyte continuum); `uncertain` = a progenitor best state without a clear margin.

```{r calls}
d <- cbind(cells, res); lv <- unique(d$true_population)
calls <- c("progenitor_in", "progenitor_review", "uncertain", "monocyte_precursor", "mature_mnp", "not_progenitor", "not_scored")
tab <- table(factor(d$true_population, lv), factor(d$progenitor_call, calls)); tab
```

```{r calls-plot}
long <- as.data.frame(tab); names(long) <- c("population", "call", "n"); long$frac <- ave(long$n, long$population, FUN = function(x) x / sum(x))
p1 <- ggplot(long, aes(call, factor(population, rev(lv)), fill = frac)) + geom_tile() + geom_text(aes(label = ifelse(n > 0, n, ""), colour = frac > 0.5), size = 3) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b", limits = c(0, 1), name = "fraction") + scale_colour_manual(values = c("black", "white"), guide = "none") +
  labs(x = NULL, y = NULL, title = "Calls by true population (cell counts)") + theme_minimal() + theme(axis.text.x = element_text(angle = 40, hjust = 1))
p1
```

```{r scatter}
s <- d[d$qc_pass, ]
ggplot(s, aes(margin, top1, colour = true_group)) + geom_point(size = 0.9, alpha = 0.6) + geom_vline(xintercept = 100, linetype = 2) + geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "margin (best minus runner-up decision value)", y = "top1 (best decision value)", colour = "true group", title = "Dashed lines: default thresholds (margin >= 100, top1 >= 0)") + theme_minimal()
```

Transitional populations (Early GMP, GMP-Mono, Pre-cDC, Pre-pDC) are blended with a neighbour state, so many land in `uncertain` rather than `progenitor_in`: **a call needs a clear lead over the runner-up (`margin`) and a good match to the state (`top1`)**. Cells that match no state, such as the simulated out-of-reference population, have very negative `top1`. Use `margin` and `top1`, not probabilities (they are saturated for this model).

## 5. Thresholds, and cases to know about

The default call needs `margin >= 100` **and** `top1 >= 0`. The `top1` gate is the expensive part in marrow-like data: on 9 held-out real bone-marrow donors (`docs/validation/validation_in_progenitor_margin_top1.csv`) in-progenitor recall at `margin >= 100` is 0.62 (precision 0.89) without it, 0.58 (0.91) with `top1 >= -500` and 0.34 (0.93) with `top1 >= 0`. It is also the guard against forcing non-marrow cells onto progenitor states: on a 210,839-cell non-marrow myeloid test set (macrophage, monocyte and DC cells from tissues), which holds essentially no progenitors, `margin >= 100` alone gives 962 `progenitor_in` calls, `top1 >= -500` gives 153 and `top1 >= 0` gives 12. Keep the default for tissue data; for marrow-like input you can lower it.

```{r sweep}
s <- d[d$qc_pass, ]; truth_in <- s$true_group %in% c("myeloid_progenitor", "dc_precursor"); g_in <- s$bm_group %in% c("myeloid_progenitor", "dc_precursor")
sweep <- do.call(rbind, lapply(seq(0, 800, 20), function(t) {
  mar <- g_in & s$margin >= t; both <- mar & s$top1 >= 0
  data.frame(margin_threshold = t, recall_margin_only = sum(mar & truth_in) / sum(truth_in), recall_margin_and_top1 = sum(both & truth_in) / sum(truth_in),
             false_calls_margin_only = sum(mar & !truth_in), false_calls_margin_and_top1 = sum(both & !truth_in))
}))
ggplot(sweep, aes(margin_threshold)) + geom_line(aes(y = recall_margin_only, linetype = "margin only")) + geom_line(aes(y = recall_margin_and_top1, linetype = "margin and top1 >= 0 (default)")) +
  geom_vline(xintercept = 100, linetype = 3) + ylim(0, 1) + labs(x = "minimum margin", y = "recall of simulated GMP / DC precursors", linetype = NULL, title = "Threshold trade-off (simulated data)") + theme_minimal()
sweep[sweep$margin_threshold %in% c(0, 100, 200, 400), ]
```

```{r cases}
oor <- d[d$true_population == "Not in reference" & d$qc_pass, ]
cat("Not-in-reference cells forced onto bone-marrow states:\n"); print(head(sort(table(oor$bm_state), decreasing = TRUE), 5))
cat("Called progenitor_in or progenitor_review:", sum(oor$progenitor_call %in% c("progenitor_in", "progenitor_review")), "of", nrow(oor), "\n")
cat("QC-failed cells scored:", sum(d$progenitor_call != "not_scored" & !d$qc_pass), "\n")
bad <- counts; rownames(bad) <- sprintf("ENSG%011d", seq_len(nrow(bad)))
cat("Ensembl-style row names ->", tryCatch(bmps_score(bad, quiet = TRUE), error = function(e) conditionMessage(e)), "\n")
```

The tool is a closed-set classifier, so out-of-reference cells are forced onto the nearest bone-marrow state; outside marrow-like tissue, `not_progenitor` is not evidence about lineage. Low-depth cells (below 500 UMIs here) are skipped when you pass a QC flag. Wrong gene identifiers stop with an error.

## 6. Checks, and agreement with the Python notebook

```{r checks}
py <- read.csv("data/sim_scores_python.csv", row.names = 1, na.strings = c("", "NA"))
ok <- !is.na(py$bm_state)
stopifnot(identical(rownames(py), res |> rownames()),
          identical(res$progenitor_call, py$progenitor_call), all(res$bm_state[ok] == py$bm_state[ok]), all(res$runner_up[ok] == py$runner_up[ok]),
          max(abs(res$top1[ok] - py$top1[ok])) < 1e-2, max(abs(res$margin[ok] - py$margin[ok])) < 1e-2)
cat(sprintf("R == Python: %d cells identical calls and states; max |top1 diff| %.1e, max |margin diff| %.1e\n", nrow(py),
            max(abs(res$top1[ok] - py$top1[ok])), max(abs(res$margin[ok] - py$margin[ok]))))
stopifnot((res$progenitor_call == "not_scored") == !cells$qc_pass,
          !any(oor$progenitor_call %in% c("progenitor_in", "progenitor_review")),
          mean(d$progenitor_call[d$true_population == "HSC" & d$qc_pass] %in% c("progenitor_review", "uncertain")) >= 0.9)
cat("all checks passed\n")
```

## 7. Cite

```{r cite}
bmps_citation()
```
